---
title: "Unsupervised learning"
subtitle: "Independent practice"
author: ""
date: ""
output: html_notebook
editor_options: 
  markdown: 
    wrap: sentence
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(message = FALSE,warning = FALSE)
```

In this guided practice, you will make clusters based on information in the `wb_bank_data_2019.csv` with information constructed by the World Bank on the distribution of the employed population by major sectors of activity and occupational category:

-   SL.AGR.EMPL.ZS: total % employment in agriculture: consisting of agriculture, forestry, hunting and fishing activities.

-   SL.IND.EMPL.ZS: % total employment in industry: in mining and quarrying, manufacturing, construction, and energy, gas and water activities.

-   SL.SERV.EMPL.ZS % total employment in services comprises wholesale and retail trade and restaurants and hotels, transportation, storage and communications, finance, insurance, real estate and business services, and personal, social and community services.

-   SL.FAM.WORK.ZS: Total % of family workers.

-   SL.EMP.MPYR.ZS: % total employers

-   SL.EMP.SELF.ZS: % total self-employed (sum of employers and own account and family workers)

-   SL.EMP.WORK.ZS: % total salaried workers

1.  Import the dataset and libraries.

```{r}

```

2.  Complete the blanks of the following code, selecting the variables. Run the code and describe what you see in the correlation plot.

```{r}
cor <- df %>% 
  select(     ) %>% 
  cor()

corrplot::corrplot(cor)
```

3.  Create the preprocessing of the recipe for a Hierarchical Clustering model. What steps are necessary to perform here? Should we normalize the variables?

```{r}

```

4.  Build the euclidean distance matrix for the data.

```{r}

```

5.  Build three hierarchical clustering models, trying different kinds of linkage: average, complete and single.

```{r}


```

6.  Evaluate the clusterings with dendograms. Which one outperformed the others?

```{r}

```

7.  Select one of the methods, define a number of clusters and describe the observations within said clusters.

```{r}

```


