---
title: "Word embeddings"
author: ''
date: ''
output:
  html_document:
    df_print: paged
subtitle: Independent practice
bibliography: references.bib
---

```{r setup}
library(tidyverse)
library(word2vec)
```

In this practice we are going to search some more cultural stereotypes.

## Loading word embedding (Word2Vec) and dataset
First, let's load the word2vec embedding model: 
```{r}
w2v <- read.word2vec("./models/GoogleNews-vectors-negative300.bin",
                     normalize = TRUE)
```

Load the affluence words:
```{r}
affluence <- read_csv('./data/affluence_words.csv')
```

## Poverty stereotypes
Let's recreate the `affluence` vector. First, we extract and aggregate the `rich` embeddings:
```{r}
```
Now, we create the `poor` pole
```{r}
```

### Food and poverty
¿Which kind of food is more related to affluence? First, let's get the `food` embedding
```{r}
```

Now, let's substract to that vector, the `poor` vector and add the `rich` vector.
```{r}
```

Finally, we will do the inverse operation.
```{r}
```
If you check the top 20 words it is possible to see that "organic food", "gourmet cousine, "nutritious food", etc. are closer to the `rich` vector. 


### Sports and poverty
Which sports are more related to each pole of the `affluence` vector?
Get the `sport` vector:
```{r}
```

Find the closest sports to the `poor` vector 
```{r}
```

Find the closests sports to the `rich` vector:
```{r}
```