---
title: "Introduction I"
subtitle: "Independent practice"
author: ""
date: ""
output: html_notebook
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(message = FALSE,warning = FALSE)
```

```{r}
library(tidyverse)
library(janitor)
library(openxlsx)
```


## Import the data set

1. Load the summary_data set.RDS data set:

```{r}

```

2. Transform the data set column names to make your workflow easier:

```{r}

```

## Explore and transform

3. Keep only rows with values greater than 70 in 'obs'. Remove 'obs' column.

```{r}

```

4. Generate a new column called 'f_age_grouped' that converts the existing 'age_grouped' column from a character vector to a factor vector. Ensure that the levels of the factor are ordered from youngest to oldest. Remove the original 'age_grouped' column.

```{r}

```

5. Order rows according to sex and race.

```{r}

```

6. Keep rows where average income is greater then 60,000 or smaller than 55,000.

```{r}

```

7. Order rows according to average income in ascending order and analyze the relationship between age group and income. Can you point out any particularly interesting or noteworthy cases?

```{r}

```

## Export

8. Export your results as an excel file called "my_independent_analysis".

```{r}

```

