---
title: "Linear Regression"
subtitle: "Challenge"
author: ""
date: ""
output: html_notebook
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(message = FALSE,warning = FALSE)
```

For this challenge, you will explore a new database. We will be using data published by [PROPUBLICA](https://www.propublica.org/article/machine-bias-risk-assessments-in-criminal-sentencing) analyzing the COMPAS Recidivism Algorithm, an algorithm designed to assess a criminal defendant's likelihood of becoming a recidivist.

```{r}
library(tidyverse)
df <- read.csv("data/machine_bias/compas-scores-two-years.csv") %>% 
        filter(score_text != 'N/A') %>% 
        select(-ends_with(".1"))
```

1.  Explore the data set and build a simple linear regression using `decile_score` as outcome variable. Explore different models using different predictor variables and interpret your results.
2.  Build a multiple linear regression using `decile_score` as outcome variable. Explore different models using different predictor variables and interpret your results.
