---
title: "Text Mining"
subtitle: "Independent practice"
author: ""
date: ""
output: html_notebook
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(message = FALSE,warning = FALSE)
```

## Text pre processing

For this independent practice, you will process the text of scraped biographies of feminists in Wikipedia.

1. Import the libraries `tidyverse`, `tidytext`, `textstem` and read the file "feminists_wiki.csv" contained in the folder "data".

```{r}

```

2. Convert the text to lower case. 

```{r}

```

3. Remove digits, punctuation signs and rare characters.

```{r}

```

4. Tokenize the text into unigrams. 

```{r}

```

5. Lemmatize the tokens.  

```{r}

```

6. What are the 10 most common words? Display them in a plot and explore them in the data frame. 

```{r}

```

7. Remove general and corpus-specific stop words. 

```{r}

```

8. Identify the new 10 most common words. 

```{r}

```
10. Create the tf-idf values for each word. 

```{r}

```

11. Explore the terms with highest TF-IDF values for two different biographies. From these tokens, what can you infer from the different figures? 

```{r}

```

