---
title: "Tidyverse + Dataviz"
subtitle: "Independent practice"
author: ""
date: ""
output: html_notebook
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(message = FALSE,warning = FALSE)
```

# COVID-19 in USA

In this independent practice, you will use the tools presented in class to process, transform, summarize and visualize information using `tidyverse`.

First, we will load the libraries and the datasets. We will use the datasets `us_covid19_daily.csv`, `us_counties_covid19_daily.csv` and `acs2017_county_data.csv`. To answer the questions, look at the information regarding the [data definitions](https://covidtracking.com/about-data/data-definitions) for the dataframes.

```{r}

```

1.  Which state has the most counties?

```{r}

```

2.  What are the 5 counties with highest count of COVID-related deaths?

```{r}

```

3.  What was the month of the year with highest count of people hospitalized? And in ICU?

```{r}

```

4.  Convert the columns of the COVID dataset to their correct format, and express the columns of numeric values in thousands.

```{r}

```

5.  What is the state with highest number of deaths per capita (in 2017 population)?

```{r}

```
