---
title: "Social Network Analysis"
subtitle: "Independent practice"
author: ""
date: ""
output: html_notebook
---

Load the libraries needed

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(ggraph)
library(igraph) 
```

## CORA Dataset

We are going to use the CORA dataset: You can see the original article at:

<https://link.springer.com/article/10.1023/A:1009953814988>

And we are taking the data originally from: <https://graphsandnetworks.com/the-cora-dataset/>

1.  Load the data from the folder 'data/cora/' into the dataframes `raw_nodes` and `raw_edges`.

```{r}

```

2.  Check the format of nodes and links of the dataframes

```{r}

```

```{r}

```

3.  Select variables

-   The `raw_nodes` data frame contains the attributes of each node. We only want to keep the `nodeId` and `subject`.
-   The `raw_edges` data frame has the links, with the clarification that the type of relation is 'cites', so the source node cites the target node, this is a directed network. We only want to keep the `sourceNodeId` and `targetNodeId`.

```{r}

```

## Build the igraph object

4.  Use the edgelist to build the graph `g` using `graph_from_data_frame`. We want to add the nodes attributes, so we can pass the `raw_nodes` dataframe to the `vertices` parameter.

```{r}

```

## Inspection

5.  Take a look at the graph. How many nodes and links are there?

```{r}

```

6.  Check the subject of the first 10 nodes.

```{r}

```

## Node level metrics

7.  Build a dataframe with the name, subject and centrality measures of nodes, and call it `nodes_df`.

```{r}

```

Which node has the highest degree?

## Describe subjects centrality

8.  Use `nodes_df` to find subject level average centrality with `group_by` and `summarise`.

```{r}

```

-   How many nodes are there per subject?
-   Which subject has the highest degree?
-   Which subject has the highest betweenness? Can you imagine why?

## Network level metrics

9.  Check the density of the network

```{r}

```

10. Is our network connected?

```{r}

```

11. If not, how many distintct components are?

```{r}

```

12. Which is the size of the components?

```{r}

```

13. Which is the diameter of the network?

```{r}

```

14. Which is the diameter of the network? Does it follow the 6-degrees of separation rule?

```{r}

```

15. Which is the **clustering coefficient** of the network?

```{r}

```

-   Is this network as clustered as the collaboration network from the guided practice? can you give a rationale of why?

## Network visualization

16. To simplify the exercise, build a the subgraph with all components that have more than 4 nodes, but less than 1000.

```{r}

```

17. Now visualize the subgraph, and color the nodes by their subject

```{r}

```

-   Are papers from the same subject connected?

18. Now use the `decompose` function to keep the giant component and visualize it

```{r}

```

-   Do you see any pattern by subject?
